Convective heat transport in stratified atmospheres at low and high Mach number

### 2017-06-13 tex/nusselt_paper.tex
lots of small tweaks

\begin{abstract}
%Convection in astrophysical systems is stratified and
%often occurs at high Rayleigh number (Ra) and low
%Mach number (Ma).
We study stratified convection in the context of 
plane-parallel, polytropically stratified atmospheres. 
We perform a suite of 2D and 3D simulations in which we vary the initial
superadiabaticity ($\epsilon$) and the Rayleigh number (Ra) at a set
initial density stratification, aspect
ratio, and Prandtl number.
We find that the Mach number (Ma) is a strong function of $\epsilon$
in both 2D and 3D simulations, and that it is a 
weak function of Ra in 2D.
Surprisingly, we find that the evolved heat transport, 
quantified by the Nusselt number (Nu),
follows scaling relationships similar to those found in the well-studied, 
incompressible Rayleigh-B\'{e}nard problem.
The scaling of evolved fluid properties (Nu, Reynolds number,
stratification) differs in the subsonic and supersonic regimes; 
aside from the sonic transition, the evolved fluid properties
of low and high Ma flows are nearly identical.  
Our limited 3D runs suggest that
evolved fluid properties behave similarly in 2D and 3D.
\end{abstract}
\maketitle

@@

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
We examine a monatomic ideal gas with an adiabatic index of
$\gamma = 5/3$ whose equation of state is $P = R\rho T$. This is consistent with the approach used in earlier work \cite{graham1975, chan&all1982,
hurlburt&all1984, cattaneo&all1990, cattaneo&all1991, brummell&all1996} 
and is the simplest stratified extension of RB.
In this work we take $R = 1$.
We study atmospheres which are initially polytropically stratified,
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(1 + L_z - z)^m, \\
T_0(z)    &= T_{t}(1 + L_z - z),
\label{eqn:polytrope}
\end{split}
\end{equation}
where $m$ is the polytropic index and $L_z$ is the depth of the atmosphere.
The height coordinate, $z$, increases upwards in the range $[0, L_z]$.
Subscript 0 indicates initial conditions and subscript $t$ indicates values
at $z = L_z$.   We
specify the depth of the atmosphere, $L_z = e^{n_{\rho}/m} - 1$, by choosing
the number of density scale heights, $n_{\rho}$, it spans initially.
Throughout this letter we set $n_{\rho} = 3$.  Satisfying hydrostatic
equilibrium sets the value of gravity, $g = T_t (m + 1)$, which is
constant with depth.  We study atmospheres with aspect
ratios of 4 where both the $x$ and $y$ coordinates have the range $[0, 4L_z]$.
In our 2D cases, we only consider $x$ and $z$.

Convective dynamics are controlled by the superadiabaticity
of the atmospheres as well as the atmospheric diffusivities.
The superadiabaticity, or the magnitude of the (negative) 
entropy gradient, is set by 
the superadiabatic excess, $\epsilon = m_{ad} - m$, where 
$m_{ad} = (\gamma - 1)^{-1}$ and the initial entropy
difference between the top and bottom boundaries is
$\Delta S_0 = \epsilon\ln (1 + L_z)$.  At a fixed value of
$\epsilon$, the diffusivities are set by the
Rayleigh number (Ra) and the Prandtl number (Pr),
\begin{equation}
\text{Ra}_{t} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu_t\chi_t},
\qquad
\text{Pr} = \frac{\nu}{\chi},
\end{equation}
where $c_P = \gamma/(\gamma-1)$ is the specific heat at
constant pressure, and $\nu$ and $\chi$ are the thermal diffusivity and
kinematic viscosity, respectively.
The atmospheric thermal diffusivity, $\chi$,
and kinematic viscosity (viscous diffusivity), $\nu$,
are determined by the Rayleigh number (Ra) and the Prandtl
number (Pr).  These polytropes are initially in
thermal equilibrium, so the initial thermal
conductivity, $\kappa_0 = \chi \rho_0$ must be
constant.  Throughout this work we specify
that Pr $= 1$ and is constant with depth.  By this
choice, $\chi = \chi_t / \rho_0$ and $\nu = \nu_t / \rho_0$.
We leave $\kappa = \rho\chi$ and the dynamic viscosity,
$\mu = \rho\nu$, free to evolve as the density 
profile evolves.  This formulation 
means Ra at the bottom of the domain is greater than
Ra$_t$ by a factor of $e^{2n_\rho}$, and henceforth
when we specify Ra we are referring to Ra$_t$.

Our diffusivities thus scale as
\begin{equation}
\chi_t \propto \sqrt{\frac{\epsilon}{\text{Pr}\cdot\text{Ra}_t}}, \qquad
\nu_t \propto \sqrt{\frac{\epsilon\cdot\text{Pr}}{\text{Ra}_t}},
\end{equation}
and for constant Pr = 1, as in this work, $\chi_t = \nu_t$.
We carry out two experiments in this letter: one in which
we fix $\epsilon$ and vary Ra, thus increasing the diffusive
timescales, and a second in which we fix Ra and vary $\epsilon$,
thus...

@@

We impose impenetrable, stress free, fixed temperature boundary conditions at
the top and bottom of the domain such that 
$w = \partial_z u = T_1 = 0$ at $z = \{0, L_z\}$. 
$T_1$ is initially filled with
random white noise whose magnitude is infinitesimal
compared to $T_0$ and $\epsilon$.
We filter this noise spectrum in coefficient space, 
such that only the lower 25\% of the coefficients
have power. We nondimensionalize our computational domains by setting
all thermodynamic variables to unity at $z = L_z$, choosing
$T_t = \rho_t = 1$.  By this choice, the non-dimensional
length scale is the inverse temperature gradient scale and the 
timescale is the isothermal sound crossing time, 
$\tau_I$, of this unit length.
Meaningful convective dynamics occur on 
timescales of the atmospheric buoyancy time,
$t_b = \tau_I \sqrt{L_z/g\epsilon}$. 
All reported results are taken from time averages
over many $t_b$ beginning \{100, 40\}$t_b$
after the start of our \{2D, 3D\} simulations in order to
assure our results are not biased by the convective transient.

@@

We measure the adiabatic Mach number (Ma = $|\bm{u}|/\sqrt{\gamma T}$).
We find that Ma is a strong function of 
$\epsilon$ and a weak function of Ra.  
In our 2D runs, when Ma $< 1$, we observe a scaling law of
Ma(Ra$, \epsilon) \appropto \epsilon^{1/2}$Ra$^{1/4}$.
This relation breaks down as the mean
Ma approaches 1 (see Fig. \ref{fig:ma_v_eps}).  This transition
occurs near Ra/Ra$_\text{crit} \approx \{10^{2}, 10^{3}\}$ for $\epsilon = \{1, 0.5\}$.
In our limited 3D runs, Ma appears to be a function of $\epsilon$ alone, with
Ma $\appropto \epsilon^{1/2}$, so at high Ra, Ma$_{\text{3D}} < \text{Ma}_{\text{2D}}$.
The value of $\epsilon$ also sets the size of the
evolved thermodynamic fluctuations, such that
$T_1/T_0 \appropto \epsilon$ and $\rho_1/\rho_0 \appropto \epsilon$.  
These fluctuations are very small when $\epsilon$ and Ma are small,
but can be O(1) for values of $\epsilon$ near 1.

@@

Low Ma flows (e.g., $\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
At high Ma (e.g., $\epsilon = 0.5$, Ra/Ra$_{\text{crit}} \gtrsim 10^3$), 
bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
(Fig. \ref{fig:entropy_snapshots}b\&c).
These shock phenomena were reported in
two \cite{cattaneo&all1990} and 
three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As Ra is increased to large values 
(Fig. \ref{fig:entropy_snapshots}c), thermodynamic structures 
no longer span the domain depth; rather, they break up into 
small eddies which traverse the domain multiple
times before diffusing.  Our choice of fixed temperature, stress free boundary conditions 
allows the flux at the boundaries to vary, while also allowing
for mean horizontal velocity profiles along the boundaries (winds).
We find these windy convective states, reminiscent of similar states
in RB convection \cite{goluskin&all2014}, at $\epsilon = 10^{-4}$ and
Ra/Ra$_{\text{crit}} \gtrsim 10^{3}$. Roll states (Fig. \ref{fig:entropy_snapshots}a) with high convective
transport alternate with windy states of lower transport.

@@

The variation of Nu with Ra is shown in 
Fig. \ref{fig:re_and_nu_v_ra}a.  We find that the Nu depends primarily
on Ra, not on $\epsilon$, except where dynamical regimes change.
In 2D and at low to moderate Ra, 
Nu $\appropto$ Ra$^{1/3}$ regardless of $\epsilon$,
reminiscent of scaling laws in \RB boundary layer theory 
\cite{grossman&lohse2000, ahlers&all2009}.
As the flow becomes supersonic,  Nu $\appropto$ Ra$^{1/5}$.
2D runs at $\text{Ra }\gtrsim 10^3$Ra$_{\text{crit}}$
and $\epsilon = 10^{-4}$ exhibit windy convective states.
In these runs, the system is in flux equilibrium only when
time averaged over roll states and windy states.
A long time average
over both of these states produces a flat flux profile which
can be sensibly analyzed, but the presence of these time-dependent states
makes it difficult to equilibrate these higher Ra cases.
It is also important to note that
the value of Nu is heavily dependent upon the specific thermodynamic
structures of the solution, as double roll states will transport
heat more efficiently than single roll states, and slight changes in
Ra can result in a simulation latching on to one solution or the other.
In our limited 3D runs, it appears that Nu $\appropto$ Ra$^{2/7}$, a classic scaling law
seen in RB studies \cite{johnston&doering2009}.

@@

We observe interesting phenomena, such as time-dependent, windy
states of shear in our 2D simulations.  
Unlike in RB \cite{goluskin&all2014}, our solutions with
windy states have transport which scales in line with our non-windy solutions.
Furthermore, the stratification of 
these polytropic atmospheres evolves in a complex
manner.  Future work should aim to 
understand the importance of stratification on
convective heat transport and other flow properties.

>>>>

\begin{abstract}
%Convection in astrophysical systems is stratified and
%often occurs at high Rayleigh number (Ra) and low
%Mach number (Ma).
We study stratified convection in the context of 
plane-parallel, polytropically stratified atmospheres. 
We perform a suite of 2D and 3D simulations in which we vary the initial
superadiabaticity ($\epsilon$) and the Rayleigh number (Ra) while fixing the
initial density stratification, aspect
ratio, and Prandtl number.
The evolved Mach number (Ma) scales according to
$\text{Ma} \propto \epsilon^{1/2}\text{Ra}^{\alpha}$, where
$\alpha = 0$ in 3D and $\alpha \approx 1/4$ in 2D.
The evolved heat transport, 
quantified by the Nusselt number (Nu),
follows scaling relationships similar to those found in the well-studied, 
incompressible Rayleigh-B\'{e}nard problem in both 2D and 3D.
At large values of $\epsilon$, significant density inversions appear in the
evolved atmospheres and seem to persist in both 2D and 3D.
\end{abstract}
\maketitle

@@

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
We examine a monatomic ideal gas with an adiabatic index of
$\gamma = 5/3$ whose equation of state is $P = R\rho T$. This is consistent with the approach used in earlier work \cite{graham1975, chan&all1982,
hurlburt&all1984, cattaneo&all1990, cattaneo&all1991, brummell&all1996} 
and is the simplest stratified extension of RB.
We study atmospheres which are initially polytropically stratified,
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(1 + L_z - z)^m, \\
T_0(z)    &= T_{t}(1 + L_z - z),
\label{eqn:polytrope}
\end{split}
\end{equation}
where $m$ is the polytropic index and $L_z$ is the depth of the atmosphere.
The height coordinate, $z$, increases upwards in the range $[0, L_z]$.
Subscript 0 indicates initial conditions and subscript $t$ indicates values
at $z = L_z$.   We
specify the depth of the atmosphere, $L_z = e^{n_{\rho}/m} - 1$, by choosing
the number of density scale heights, $n_{\rho}$, it spans initially.
Throughout this letter we set $n_{\rho} = 3$.  Satisfying hydrostatic
equilibrium sets the value of gravity, $g = T_t (m + 1)$, which is
constant with depth.  We study atmospheres with aspect
ratios of 4 where both the $x$ and $y$ coordinates have the range $[0, 4L_z]$.
In our 2D cases, we only consider $x$ and $z$.

We nondimensionalize our domains by setting
all thermodynamic variables to unity at $z = L_z$, choosing
$R = T_t = \rho_t = 1$.  By this choice, the non-dimensional
length scale is the inverse temperature gradient scale and the 
timescale is the isothermal sound crossing time, 
$\tau_I$, of this unit length.
Meaningful convective dynamics occur on 
timescales of the atmospheric buoyancy time,
$t_b = \tau_I \sqrt{L_z/g\epsilon}$.

At fixed $n_\rho$, convective dynamics are 
controlled by the superadiabaticity
of the atmospheres as well as the atmospheric diffusivities.
The superadiabaticity, or the magnitude of the (negative) 
entropy gradient, is set by 
the superadiabatic excess, $\epsilon = m_{ad} - m$, where 
$m_{ad} = (\gamma - 1)^{-1}$ and the initial entropy
difference between the top and bottom boundaries is
$\Delta S_0 = \epsilon\ln (1 + L_z)$.  At a fixed value of
$\epsilon$, the diffusivities are set by the
Rayleigh number (Ra) and the Prandtl number (Pr),
\begin{equation}
\text{Ra}_{t} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu_t\chi_t},
\qquad
\text{Pr} = \frac{\nu}{\chi},
\end{equation}
where $c_P = \gamma/(\gamma-1)$ is the specific heat at
constant pressure, and $\nu$ and $\chi$ are the thermal diffusivity and
kinematic viscosity, respectively.
These polytropes are initially in
thermal equilibrium, so the initial thermal
conductivity, $\kappa_0 = \chi \rho_0$, must be
constant.  Throughout this work we specify
that Pr $= 1$ and is constant with depth.  By these
choices, $\chi = \chi_t / \rho_0$ and $\nu = \nu_t / \rho_0$.
This formulation 
means Ra at the bottom of the domain is greater than
Ra$_t$ by a factor of $e^{2n_\rho}$, and henceforth
when we specify Ra we are referring to Ra$_t$.  
We leave $\kappa = \rho\chi$ and 
$\mu = \rho\nu$ (the dynamic viscosity) free to evolve as the density 
profile evolves.

Atmospheric diffusivities thus scale as
\begin{equation}
\chi_t \propto \sqrt{\frac{\epsilon}{\text{Pr}\cdot\text{Ra}_t}}, \qquad
\nu_t \propto \sqrt{\frac{\epsilon\cdot\text{Pr}}{\text{Ra}_t}},
\label{eqn:diffusivities}
\end{equation}
and for constant Pr = 1, as in this work, $\chi_t = \nu_t$.
We carry out two experiments in this letter: one in which
we fix $\epsilon$ and increase Ra, thus increasing the diffusive
timescales, and a second in which we fix Ra and raise $\epsilon$,
thus decreasing the ratio of the buoyant timescale to the 
diffusive timescales.

@@

We impose impenetrable, stress free, fixed temperature boundary conditions at
the top and bottom of the domain such that 
$w = \partial_z u = T_1 = 0$ at $z = \{0, L_z\}$. 
$T_1$ is initially filled with
random white noise whose magnitude is infinitesimal
compared to $T_0$ and $\epsilon$.
We filter this noise spectrum in coefficient space, 
such that only the lower 25\% of the coefficients
have power. All reported results are taken from time averages
over many $t_b$ beginning \{100, 40\}$t_b$
after the start of our \{2D, 3D\} simulations in order to
assure our results are not biased by the convective transient.

@@

We measure the adiabatic Mach number (Ma = $|\bm{u}|/\sqrt{\gamma T}$).
We find that Ma is a strong function of 
$\epsilon$ and a weak function of Ra.  
In our 2D runs, when Ma $< 1$, we observe a scaling law of
Ma(Ra$, \epsilon) \appropto \epsilon^{1/2}$Ra$^{1/4}$.
This relation breaks down as the mean
Ma approaches 1 (see Fig. \ref{fig:ma_v_eps}).  This transition
occurs near Ra/Ra$_\text{crit} \approx \{10^{2}, 10^{3}\}$ for $\epsilon = \{1, 0.5\}$.
In our limited 3D runs, Ma appears to be a function of $\epsilon$ alone, with
Ma $\appropto \epsilon^{1/2}$, so at high Ra, Ma$_{\text{3D}} < \text{Ma}_{\text{2D}}$.

@@

Low Ma flows (e.g., $\epsilon = 10^{-4}$)
display the classic narrow downflow and broad upflow lanes of stratified
convection (Fig. \ref{fig:entropy_snapshots}a).
At high Ma (e.g., $\epsilon = 0.5$, Ra/Ra$_{\text{crit}} \gtrsim 10^3$), 
bulk thermodynamic structures are similar but
shock systems form in the upper atmosphere near downflow lanes 
(Fig. \ref{fig:entropy_snapshots}b\&c).
These shock phenomena were reported in
two \cite{cattaneo&all1990} and 
three \cite{malagoli&all1990} dimensional polytropic simulations previously.
As Ra is increased to large values (Fig. \ref{fig:entropy_snapshots}c) 
and diffusivities become small according to (\ref{eqn:diffusivities}),
thermodynamic structures break up into small eddies which traverse the
domain repeatedly before diffusing.
Furthermore, the value of $\epsilon$ sets the size of the
evolved thermodynamic fluctuations from their adiabatic values,
as evidenced by the scalings of the colorbars.
%These fluctuations are very small when $\epsilon$ and Ma are small,
%but can be O(1) for values of $\epsilon$ near 1.

%Our choice of fixed temperature, stress free boundary conditions 
%allows the flux at the boundaries to vary, while also allowing
%for mean horizontal velocity profiles along the boundaries (winds).
%We find these windy convective states, reminiscent of similar states
%in RB convection \cite{goluskin&all2014}, at $\epsilon = 10^{-4}$ and
%Ra/Ra$_{\text{crit}} \gtrsim 10^{3}$. Roll states (Fig. \ref{fig:entropy_snapshots}a) with high convective
%transport alternate with windy states of lower transport.

@@

The variation of Nu with Ra is shown in 
Fig. \ref{fig:re_and_nu_v_ra}a.  We find that the Nu depends primarily
on Ra, not on $\epsilon$, except where dynamical regimes change.
In 2D and at low to moderate Ra, 
Nu $\appropto$ Ra$^{1/3}$ regardless of $\epsilon$,
reminiscent of scaling laws in \RB boundary layer theory 
\cite{grossman&lohse2000, ahlers&all2009}.
As the flow becomes supersonic,  Nu $\appropto$ Ra$^{1/5}$.
%2D runs at $\text{Ra }\gtrsim 10^3$Ra$_{\text{crit}}$
%and $\epsilon = 10^{-4}$ exhibit windy convective states.
%In these runs, the system is in flux equilibrium only when
%time averaged over roll states and windy states.
%A long time average
%over both of these states produces a flat flux profile which
%can be sensibly analyzed, but the presence of these time-dependent states
%makes it difficult to equilibrate these higher Ra cases.
It is also important to note that
the value of Nu is heavily dependent upon the specific thermodynamic
structures of the solution, as double roll states will transport
heat more efficiently than single roll states, and slight changes in
Ra can result in a simulation latching on to one solution or the other. 
(ADD SOMETHING ABOUT THIS BEING ASPECT RATIO INDEPENDENT AND MAYBE BRIEFLY
MENTION WINDY STATES HERE?)
In our limited 3D runs, it appears that Nu $\appropto$ Ra$^{2/7}$, a classic scaling law
seen in RB studies \cite{johnston&doering2009}.

@@

%We observe interesting phenomena, such as time-dependent, windy
%states of shear in our 2D simulations.  
%Unlike in RB \cite{goluskin&all2014}, our solutions with
%windy states have transport which scales in line with our non-windy solutions.
The stratification of 
these polytropic atmospheres evolves in a complex
manner.  Future work should aim to 
understand the importance of stratification on
convective heat transport and other flow properties.
